---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
library(readr)
library(tidyverse)
library(rmarkdown)
```



# loading data files into R 
```{r}
# loading data files into R 
  activity <- read_csv("dailyActivity_merged.csv")
  calories <- read_csv("hourlyCalories_merged.csv")
  intensities <- read_csv("hourlyIntensities_merged.csv")
  sleep <- read_csv("sleepDay_merged.csv")
  weight <- read_csv("weightLogInfo_merged.csv")
```

 head(activity)
  # Identify all the columns in the daily_activity data.
```{r}
colnames(activity)
```
  
  
  # Take a look at the sleep_day data.
```{r}
 head(sleep)
```
  
 
  # Identify all the columns in the daily_activity data.
```{r}
colnames(sleep)
```

```{r}

  head(calories)
  head(intensities)
  head(weight)
```
  
  
  # checking data structure 
  str(activity)
  str(sleep)
  str(calories)
  str(intensities)
  str(weight)
  #Checking for duplicate rows
